\large Neural and Symbolic Optimization of Cosmological Particle-Mesh Simulation \vspace{0.3em}

### 2024-02-26 docs/papers/sto/so.tex
FIXUP draft update adjoint paper bib
FIXUP draft updates

\section*{TODO}
\begin{itemize}

@@

  example}).\newline
  \YZ{Do we also use another subset of the training data as validation data?}

@@

\item (test) Generate a sample of $(X, y)$ data with the trained neural network
  for feature selection and symbolic regression.
\item \YZ{Do we also consider pruning and distilling of the trained networks?}

@@

\begin{table}
  \centering
  \caption{Possible features that can affect force sharpening.
  All scales are then multiplied by the wavenumber $k$ to form dimensionless
  factors.}
  \label{tab:feat}
  \begin{tabular}{rl}
  \toprule
  type & feature \\
  \midrule
  & $R_\mathrm{d}$: RMS linear displacement \\
  \cmidrule(lr){2-2}
  & $R_P$: $\Delta_\linear^2(R_P^{-1}) \triangleq 1$ \\
  \cmidrule(lr){2-2}
  nonlinear scales & $R_\tophat$: $\sigma_\tophat^2(R_\tophat) \triangleq 1$ \\
  \cmidrule(lr){2-2}
  & $R_\gauss$: $\sigma_\gauss^2(R_\gauss) \triangleq 1$ \\
  \cmidrule(lr){2-2}
  & $\d R / \d\lna$: time derivatives of the above \\
  \midrule
  & particle spacing \\
  \cmidrule(lr){2-2}
  other scales & $l$: PM force mesh cell size \\
  \cmidrule(lr){2-2}
  & softening length \\
  \midrule
  & $G_m(a) \triangleq D_m(a) / a^m, m \in \{1, 2, \YL{3's?}\}$ \\
  \cmidrule(lr){2-2}
  & $\d\ln G_m / \d\lna$ \\
  \cmidrule(lr){2-2}
  dimensionless factors & $\Omegam(a), \YL{\OmegaK(a)}$ \\
  \cmidrule(lr){2-2}
  & $\d\ln\!H / \d\lna$ \\
  \cmidrule(lr){2-2}
  & $\Delta\lna$: time step size in $\lna$\YL{. Now this can be an important feature with varying $a_\ic$}\\
  \bottomrule
  \end{tabular}
  \end{table}

In cosmological $N$-body simulations, the dynamics of the particles is
governed by the Poisson equation
%
\begin{equation}
  \nabla^2 \varphi(\vx) = \frac32 \Omegam H_0^2 \delta(\vx),
\end{equation}
%
where $\Omegam$ is the matter fraction today, $H_0$ is the Hubble
constant, and the two fields $\varphi(\vx)$ and $\delta(\vx)$ are the
gravitational potential and matter density fluctuation respectively.
In Fourier space, the gravitational force field $- \vnabla \varphi$
follows
%
\begin{equation}
- i \vk \varphi(\vk) = i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
\end{equation}
%
where for each dimension $i\in\{1,2,3\}$ of the 3D field, we then
introduce SO force sharpening through
%
\begin{equation}
\frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
  f(k_i; \vvartheta) g(k_1, k_2, k_3; \vvartheta),
\end{equation}
%
where $f$ and $g$ are NNs, each approximating a nonlinear function of
the wavevector components and all possible relevant features
$\vvartheta$ summarized in \autoref{tab:feat} with more details
discussed in \autoref{sec:features}.
Gravity obeys the $\mathrm{O}(3)$ rotational symmetry, which is however
broken by the PM solver to the O\textsubscript{h} discrete symmetry with
48 elements involving permutations and reversals of the 3 principal axes
of the simulation box.
The PM evaluation of each force component can further break the symmetry
along that axis, leaving the remaining 2 permutationally symmetric.
Therefore, we implement $f$ and $g$ to be even functions of the
wavevector components, and $g$ to be permutation equivariant in all 3
components, so that $f g$ is the most general modification to the force
kernel.
An even and real kernel in configuration space is also even in Fourier
space.
We take the absolute value of the wavevectors for the evenness, and sort
the 3 wavevector compontents to achieve the permutation equivariance.

\begin{table*}
  \centering
  \caption{Ranges of \GADGET\ and \pmwd\ configuration and cosmological
  parameters.
  Note that the grid ratio need next fast len to determine the mesh shape
  for fast FFT.
  Given the box size, the mesh shape determines the cell size.
  The softening parameter gives the ratio of the comoving softening length
  to the mean particle spacing.
  The curvature $\OmegaK$ is related to the separate universe simulation.
  We sample parameters applicable to \GADGET\ during data generation, and
  those applicable to \pmwd\ during training.
  \YL{(I forgot why I wanted to sample mesh shape and mesh cell size...
  Now it seems more natural to me to have the ptcl spacing sampled
  uniformly instead...)}
  }
  \label{tab:param}
  \begin{tabular}{lcccr}
  \toprule
  applicability & parameter & distribution & range & sampling \\
  \midrule
  & $128^3$ white noise & $\mathcal{N}(\vzero, \vI)$ & 512 realizations & MC \\
  \cmidrule(lr){2-5}
  & box size in Mpc$/h$ & \YL{log-uniform} & [25.6, 2560) \\
  \cmidrule(lr){2-4}
  & snapshot offset & uniform & (0, 1] \\
  \cmidrule(lr){2-4}
  \GADGET\ & $\As \times 10^9$ & log-uniform & [1, 4) \\
  \cmidrule(lr){2-4}
  \& \pmwd\ & $\ns$ & log-uniform & [0.75, 1.25) & RQMC \\
  \cmidrule(lr){2-4}
  & $\Omegam$ & log-uniform & [1/5, 1/2) & see \autoref{fig:sobol} \\
  \cmidrule(lr){2-4}
  & $\Omegab / \Omegam$ & log-uniform & [1/8, 1/4) \\
  \cmidrule(lr){2-4}
  & $\OmegaK / (1 - \OmegaK)$ & uniform & [-1/3, 1/3) \\
  \cmidrule(lr){2-4}
  & $h$ & log-uniform & [0.5, 1) \\
  \cmidrule(lr){1-4}
  \GADGET\ & softening ratio & log-uniform & [1/50, 1/20) \\
  \cmidrule(lr){1-5}
  \pmwd\ & mesh shape & log-uniform & [128, 512] & MC \\
  \cmidrule(lr){2-4}
  \cmidrule(lr){1-4}
  training & loss grid offset & uniform & $[0, 1)^3$ \\
  \bottomrule
  \end{tabular}
  \end{table*}

Our target data samples for training and testing are generated with the
\GADGET\ simulation code \citep{GADGET-4}.
We run 512 \GADGET\ simulations that cover a wide range of
configurations and cosmological parameters, as listed in
\autoref{tab:param}.
\YL{
Each simulation evolves $128^3$ particles from the initial time $a_\ic$
to the final time $a_\f=1$, saving 64 snapshots linearly spaced in scale
factor from $a_\ic + \gamma \Delta a$ to $1 - (1 - \gamma) \Delta a$.
We determine $a_\ic$ for each configuration depending on its cosmology
and mean particle spacing \citep{MichauxEtAl2021}, \YL{(YL elaborate why and how here.)} which then sets the
snapshot spacing $\Delta a = (1 - a_\ic) / 64$.
To sample different alignment between the snapshots and the start/stop
times and avoid possible related numerical errors in \pmwd, we introduce
a fractional random offset $\gamma$ that varies between 0 and 1 for
different simulations.
}
In our \pmwd\ model, we generate the snapshot at an arbitrary given time
using the cubic Hermite interpolation, with more details in
\autoref{sec:snapobs}.
The curvature $\OmegaK$ is related to the separate universe simulation
\citep{LiEtAl2014, WagnerEtAl2015}.
\YZ{We refer to this training dataset of 512 \GADGET\ simulations sampled with
Sobol sequence as \GSDATA.}
On average each \GADGET\ simulation takes roughly 2K CPU hours to run, and the
generation of the whole \GSDATA\ dataset consumes 1M CPU hours.
% to be compare with the speed of pmwd with SO

\YL{
In this work, we train \pmwd\ using $XXX$ time steps from $a_\ic$ to
$a_\f$, generate snapshots at the same times as in \GADGET, and compare
them in the loss function.
Although the number of time steps are fixed here, we emphasize that our
spatial optimization of course also applies for different number of time
steps, which we will explore, together with the optimal choice of time
variable, in a follow-up work on the temporal optimization.
}

For both $f$ and $g$ NNs, we adopt multilayer perceptron (MLP, i.e.
fully connected) structures.
Both $f$ and $g$ networks have $7$ hidden layers with $4\times n_{\rm
input}$ nodes for each layer, where $n_{\rm input}$ is the size of the
input layer, which is [...] and [...] for $f$ and $g$ respectively.
A rectified linear unit (ReLU) activation function is used for hidden
layers.
We further apply an exponential function to the output as a regulator
for positive output.
Being embedded in the differentiable simulation library \pmwd, the NNs
can be trained using gradient-based optimization methods as in usual
deep learning tasks.
The parameters of the NNs, which enter the force evaluation in every
time integration step, can be treated just the same as other
cosmological parameters.
The objective gradients are accumulated, based on which the parameters
are then optimized.

With discrete Fourier transform (DFT) on a regular mesh of length $N_c$
with cell size $l_c$, the range of sampled 1D discrete wavenumbers is
given by the fundamental $k_{\rm fund} = 2\pi/(l_c N_c)$ and the
maximum $k_{\rm Nyquist} = \pi/l_c$, where $l_c N_c$ is simply the box
size.
To visualize the functions for different box sizes, we normzalize $k$ by
$k_{\rm Nyquist}$, which gives the same $[2/N_c, 1]$ regardless of the
box size.
For the 3D wavenumber $k=\sqrt{\sum_{i=1}^{3}k_i^2}$, the maximum value
is then $\sqrt{3}$.

For the $g(k_1,k_2,k_3;\vvartheta)$, which is permutationally
symmetric about $(k_1, k_2, k_3)$, we show it as a function of $k_i$
along axis, face and body diagonal directions.

\subsection{Analytic Expressions with Symbolic Regression}

We fit the trained NNs with symbolic regression.

To compare two fields $\hat f$ and $f$, we introduce for each mode the
square root ratio of the auto power spectrum
%
\begin{equation}
T(\vk) \triangleq
\sqrt{\frac{|\hat f(\vk)|^2}{|f(\vk)|^2}},
\label{T}
\end{equation}
%
which is referred to as the transfer function (TF), and the correlation
coefficient (CC) defined with the cross power spectrum
%
\begin{equation}
r(\vk) \triangleq
\frac{\Re \bigl[ \hat f(\vk) f^*(\vk) \bigr]}
     {\sqrt{|\hat f(\vk)|^2 |f(\vk)|^2}}.
\label{r}
\end{equation}
%
The two fields are identical when both $T$ and $r$ are unity.
The PM method is accurate on scales much larger than the mesh cell size,
therefore we expect $T$ and $r$ to be very close to being unity both
before and after the force sharpening optimization.
To show the $T$ and $r$ for different box sizes, we normalize the
wavenumber by the Nyquist wavenumber $k_{\rm Nyquist} = \pi / l_p$ of
the particle grid, where $l_p$ is the particle spacing.
\YZ{We show T and r separately for different mesh shapes?}

The spatial optimization using gradient descent discussed in this work
is built in the framework of the PM $N$-body library \pmwd\
\citep{Li2022b}, which is an implementation of the differentiable
cosmological simulation with adjoint method \citep{Li2022a}.

Here we briefly recap the equations that are closely related to our
extensions of the original framework, and introduce the notations of the
variables.
Following \cite{Li2022a}, we denote the particles' state with $\vz =
(\vx, \vp)^\intercal$, where $\vx$ and $\vp$ are the positions and
momenta of the particles, and their adjoint variables with $\vlambda =
(\vxi, \vpi)$.
We refer to the quantities that are directly related to the final
objective function $\cJ$ as the observable $\vo$.
In this work, our observables include multiple snapshots at given times.
For the observation operator $O_{i}^{i-1}$ in the adjoint equations
(check Eq.(18) in \cite{Li2022a}) from time step $a_i$ back to
$a_{i-1}$, the adjoint variables are updated through
\begin{align}
  \vxi_{i-1} &:= \vxi_{i-1} + \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p \vx_{i-1}}, \nonumber\\
  \vpi_{i-1} &:= \vpi_{i-1} + \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p \vp_{i-1}},
\end{align}
where the derivatives of $\vo$ with respect to $\vx_{i-1}$ and $\vp_{i-1}$ would
be non-zero only if they were used in the generation of any snapshots in $\vo$.
The initial conditions of these adjoint variables are given by
\begin{align}
  \vxi_n &= \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p\vx_n}, \nonumber\\
  \vpi_n &= \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p\vp_n},
\end{align}
which are zero if no snapshot in $\vo$ depends on $(\vx_n, \vp_n)$, the
state of particles at the final time step.
The derivative of the objective function with respect to the observable
($\p\cJ/\p\vo$) and the VJPs of the observation operation
($\p\vo/\p\vx_n$ and $\p\vo/\p\vp_n$, mainly the interpolation function
discussed in \autoref{sec:snapobs} below) can be conveniently computed
by AD.

The N-body time integration is carried out along discrete time steps in
terms of the scale factor $a$, and the state of particles at any given
time can be achieved using interpolation.
Given the particle displacements $\vs$ and velocities\footnote{This
velocity is related to the canonical momentum $\vp$ through $\vv =
\vp/(a^3H)$.} $\vv := \d\vs/\d a$ at two consecutive time steps $a_i$ and
$a_{i+1}$, we get the snapshot of particles in between at $a$ using the
cubic Hermite interpolation.
The interpolated particle displacements are given by
\begin{align}
  \vs(a) =\ &h_{00}(\alpha)\vs_i + h_{10}(\alpha)(a_{i+1} - a_i)\vv_i + \nonumber\\
           &h_{01}(\alpha)\vs_{i+1} + h_{11}(\alpha)(a_{i+1} - a_i)\vv_{i+1},
\end{align}
where $\alpha := (a - a_i)/(a_{i+1} - a_i)$ and the Hermite basis
functions read
\begin{align}
  h_{00}(\alpha) &= 2\alpha^3 - 3\alpha^2 + 1, \nonumber\\
  h_{10}(\alpha) &= \alpha^3 - 2\alpha^2 + \alpha, \nonumber\\
  h_{01}(\alpha) &= -2\alpha^3 + 3\alpha^2, \nonumber\\
  h_{11}(\alpha) &= \alpha^3 - \alpha^2.
\end{align}
The particle velocities are then given by the derivative $\vv(a) =
\d\vs(a) / \d a$.
The interpolation is a linear combination of the particle snapshots at
two consecutive time steps, which therefore can be done in either two
steps or just one step by carrying the particles from the previous step.
We find the two-steps scheme more convenient for the backward evolution
of the adjoint variables.
The interpolated snapshots at different times make the observable in
this work, and the VJPs of the interpolation functionx are computed by
AD and put into the adjoint equations as mentioned in
\autoref{sec:pmwd}.

For the displacement field, we adopt the mean squared error (MSE) loss
that proved effective in many previous works
\citep[e.g.,][]{HeEtAl2019, LiEtAl2021}:
%
\begin{equation}
\cL_\vs = \ln \frac{\sum_\vq \bigl[ \hat\vs(\vq) - \vs(\vq) \bigr]^2}
                   {\sum_{\vq'} s(\vq')^2},
\end{equation}
%
in which a particle originating from its Lagrangian position $\vq$ is
displaced by $\vs$ in the \GADGET\ snapshot and by $\hat\vs$ in the
prediction of \pmwd\ model in training.
Note that we have further taken the logarithm to combine it with the
other loss(es) below.
% Say that only s loss is not enough even though in principle it is, and cite DrewEtAl x2 too

Small numerical differences in initial conditions (ICs) could become
more significant with the $N$-body evolution.
We have verified that the IC generation on GPUs is deterministic though
different from those on CPUs, therefore we generate ICs for both the
training data and the model on GPUs.

We generate samples of (input, output) data of the trained neural
networks to perform symbolic regression.
We use Sobol sequence to generate the sample of (parameters in
\autoref{tab:param}, $k$, $a$).

>>>>

\begin{table*}
  \centering
  \caption{Ranges of \GADGET\ and \pmwd\ configurations and cosmological
  parameters.
  Note that the grid ratio need next fast len to determine the mesh shape
  for fast FFT.
  Given the box size, the mesh shape determines the cell size.
  The softening parameter gives the ratio of the comoving softening length
  to the mean particle spacing.
  The curvature $\OmegaK$ is related to the separate universe simulation.
  We sample parameters applicable to \GADGET\ during data generation, and
  those applicable to \pmwd\ during training.
  \YL{(I forgot why I wanted to sample mesh shape and mesh cell size...
  Now it seems more natural to me to have the ptcl spacing sampled
  uniformly instead...)}
  }
  \label{tab:param}
  \begin{tabular}{lcccr}
  \toprule
  applicability & parameter & distribution & range & sampling \\
  \midrule
  & $128^3$ white noise & $\mathcal{N}(\vzero, \vI)$ & 512 realizations & MC \\
  \cmidrule(lr){2-5}
  & box size in Mpc$/h$ & \YL{log-uniform} & [25.6, 2560) \\
  \cmidrule(lr){2-4}
  & snapshot offset & uniform & (0, 1] \\
  \cmidrule(lr){2-4}
  \GADGET\ & $\As \times 10^9$ & log-uniform & [1, 4) \\
  \cmidrule(lr){2-4}
  \& \pmwd\ & $\ns$ & log-uniform & [0.75, 1.25) & RQMC \\
  \cmidrule(lr){2-4}
  & $\Omegam$ & log-uniform & [1/5, 1/2) & see \autoref{fig:sobol} \\
  \cmidrule(lr){2-4}
  & $\Omegab / \Omegam$ & log-uniform & [1/8, 1/4) \\
  \cmidrule(lr){2-4}
  & $\OmegaK / (1 - \OmegaK)$ & uniform & [-1/3, 1/3) \\
  \cmidrule(lr){2-4}
  & $h$ & log-uniform & [0.5, 1) \\
  \cmidrule(lr){1-4}
  \GADGET\ & softening ratio & log-uniform & [1/50, 1/20) \\
  \cmidrule(lr){1-5}
  \pmwd\ & mesh shape & log-uniform & [128, 512] & MC \\
  \cmidrule(lr){2-4}
  \cmidrule(lr){1-4}
  training & loss grid offset & uniform & $[0, 1)^3$ \\
  \bottomrule
  \end{tabular}
  \end{table*}

\begin{table}
  \centering
  \caption{Possible features that can affect force sharpening.
  All scales are then multiplied by the wavenumber $k$ to form dimensionless
  factors.}
  \label{tab:feat}
  \begin{tabular}{rl}
  \toprule
  type & feature \\
  \midrule
  & $R_\mathrm{d}$: RMS linear displacement \\
  \cmidrule(lr){2-2}
  & $R_P$: $\Delta_\linear^2(R_P^{-1}) \triangleq 1$ \\
  \cmidrule(lr){2-2}
  nonlinear scales & $R_\tophat$: $\sigma_\tophat^2(R_\tophat) \triangleq 1$ \\
  \cmidrule(lr){2-2}
  & $R_\gauss$: $\sigma_\gauss^2(R_\gauss) \triangleq 1$ \\
  \cmidrule(lr){2-2}
  & $\d R / \d\lna$: time derivatives of the above \\
  \midrule
  & particle spacing \\
  \cmidrule(lr){2-2}
  other scales & $l$: PM force mesh cell size \\
  \cmidrule(lr){2-2}
  & softening length \\
  \midrule
  & $G_m(a) \triangleq D_m(a) / a^m, m \in \{1, 2, \YL{3's?}\}$ \\
  \cmidrule(lr){2-2}
  & $\d\ln G_m / \d\lna$ \\
  \cmidrule(lr){2-2}
  dimensionless factors & $\Omegam(a), \YL{\OmegaK(a)}$ \\
  \cmidrule(lr){2-2}
  & $\d\ln\!H / \d\lna$ \\
  \cmidrule(lr){2-2}
  & $\Delta\lna$: time step size in $\lna$ \\
  & \YL{Now this can be an important feature with varying $a_\ic$} \\
  \bottomrule
  \end{tabular}
  \end{table}

Cosmological $N$-body simulations evolve dark matter particles under the
gravitational force.
The dynamics of the particles is governed by the Poisson equation
%
\begin{equation}
  \nabla^2 \varphi(\vx) = \frac32 \Omegam H_0^2 \delta(\vx),
\end{equation}
%
where $\Omegam$ is the matter fraction today, $H_0$ is the Hubble
constant, and the two fields $\varphi(\vx)$ and $\delta(\vx)$ are the
gravitational potential and matter density fluctuation respectively.
In Fourier space, the gravitational force field $- \vnabla \varphi$
follows
%
\begin{equation}
- i \vk \varphi(\vk) = i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
\end{equation}
%
we then introduce SO force sharpening through
%
\begin{equation}
\frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
  f(k_i; \vvartheta) g(k_1, k_2, k_3; \vvartheta),
\end{equation}
%
where $f$ and $g$ are NNs, each approximating a nonlinear function of
the wavevector components and all possible relevant features
$\vvartheta$.
The subscript $i\in\{1,2,3\}$ denotes the 3 principal axes of the
simulation box.

Gravity obeys the $\mathrm{O}(3)$ rotational symmetry, which is however
broken by the PM solver to the O\textsubscript{h} discrete symmetry with
48 elements involving permutations and reversals of the 3 principal axes
of the simulation box.
The PM evaluation of each force component can further break the symmetry
along that axis, leaving the remaining 2 permutationally symmetric.
Therefore, we implement $f$ and $g$ to be even functions of the
wavevector components, and $g$ to be permutation equivariant in all 3
components, so that $f g$ is the most general modification to the force
kernel.
An even and real kernel in configuration space is also even in Fourier
space.
We take the absolute value of the wavevectors for the evenness, and sort
the 3 wavevector compontents to achieve the permutation equivariance.

Our target data samples are generated with the \GADGET\ simulation code
\citep{GADGET-4}.
We run 512 \GADGET\ simulations that cover a wide range of
configurations and cosmological parameters, as listed in
\autoref{tab:param}.
\YL{
Each simulation evolves $128^3$ particles from the initial time $a_\ic$
to the final time $a_\f=1$, saving 64 snapshots linearly spaced in scale
factor from $a_\ic + \gamma \Delta a$ to $1 - (1 - \gamma) \Delta a$.
We determine $a_\ic$ for each configuration depending on its cosmology
and mean particle spacing \citep{MichauxEtAl2021}, \YL{(YL elaborate why
and how here.)} which then sets the snapshot spacing $\Delta a = (1 -
a_\ic) / 64$.
To sample different alignment between the snapshots and the start/stop
times and avoid possible related numerical errors in \pmwd, we introduce
a fractional random offset $\gamma$ that varies between 0 and 1 for
different simulations.
}
The curvature $\OmegaK$ is related to the separate universe simulation
\citep{LiEtAl2014, WagnerEtAl2015}.
\YZ{We refer to this training dataset of 512 \GADGET\ simulations
sampled with Sobol sequence as \GSDATA.}
On average each \GADGET\ simulation takes roughly 2K CPU hours to run,
and the generation of the whole \GSDATA\ dataset consumes 1M CPU hours.

\YL{
In this work, we train \pmwd\ using $XXX$ time steps from $a_\ic$ to
$a_\f$, generate snapshots at the same times as in \GADGET, and compare
them in the loss function.}
In our \pmwd\ model, the snapshot at an arbitrary given time is
generated using the cubic Hermite interpolation, with more details in
\autoref{sec:snapobs}.
\YL{Although the number of time steps are fixed here, we emphasize that
our spatial optimization of course also applies for different number of
time steps, which we will explore, together with the optimal choice of
time variable, in a follow-up work on the temporal optimization.
}

For both $f$ and $g$ NNs, we adopt multilayer perceptron (MLP, i.e.
fully connected) structures to learn the nonlinear functions.
Both $f$ and $g$ networks have $XXX$ hidden layers with $XXX\times n_{\rm
input}$ nodes for each layer, where $n_{\rm input}$ is the size of the
input layer, which is [...] and [...] for $f$ and $g$ respectively.
The rectified linear unit (ReLU) activation function is used for hidden
layers.
We further apply the exponential function to the output as a regulator
for positive output.
Being embedded in the differentiable simulation library \pmwd, the NNs
can be trained using gradient-based optimization methods as in usual
deep learning tasks.
The parameters of the NNs, which enter the force evaluation in every
time integration step, can be treated just the same as other
cosmological parameters.
The objective gradients are accumulated, based on which the parameters
are then optimized.

With discrete Fourier transform (DFT) on a regular grid of length $N_c$
and cell size $l_c$, the range of sampled 1D discrete wavenumbers is
limited by the fundamental and Nyquist values, i.e. $k_{\rm fund} =
2\pi/(l_c N_c)$ and $k_{\rm Nyquist} = \pi/l_c$, where $l_c N_c$ is
simply the box size.
Then the normzalized wavenumber $k/k_{\rm Nyquist}$ gives the same range
$[2/N_c, 1]$ for different box sizes.
For the 3D wavenumber $k=\sqrt{\sum_{i=1}^{3}k_i^2}$, the maximum
normalized value is then $\sqrt{3}$.

For $g(k_1,k_2,k_3;\vvartheta)$, which is permutationally symmetric
about $(k_1, k_2, k_3)$, we show it as a function of the wavevector
along axis, face, and body diagonal directions.

\subsection{Symbolic Expressions}

We fit the trained NNs with symbolic regression for analytic
expressions of the $f(k_i; \vvartheta)$ and $g(k_1, k_2, k_3;
\vvartheta)$ functions.

Improvements of SO on the dark matter density fields are the first
things to check.
We construct these fields by scattering the particles onto a mesh.
To compare two fields $\hat f$ and $f$, we introduce for each mode the
square root ratio of the auto power spectrum
%
\begin{equation}
T(\vk) \triangleq
\sqrt{\frac{|\hat f(\vk)|^2}{|f(\vk)|^2}},
\label{T}
\end{equation}
%
which is referred to as the transfer function (TF), and the correlation
coefficient (CC) defined with the cross power spectrum
%
\begin{equation}
r(\vk) \triangleq
\frac{\Re \bigl[ \hat f(\vk) f^*(\vk) \bigr]}
     {\sqrt{|\hat f(\vk)|^2 |f(\vk)|^2}}.
\label{r}
\end{equation}
%
The two fields are identical when both $T$ and $r$ are unity.
The PM method is accurate on scales much larger than the mesh cell size,
therefore we expect $T$ and $r$ to be very close to being unity both
before and after the force sharpening optimization.
To show the $T$ and $r$ for different box sizes, we normalize the
wavenumber by the Nyquist wavenumber $k_{\rm Nyquist} = \pi / l_p$ of
the particle grid, where $l_p$ is the particle spacing.
\YZ{We show T and r separately for different mesh shapes?}

We time the computation of force sharpening for one step using NNs and
SR respectively.

The SO using gradient descent training discussed in this work is built
in the framework of the PM $N$-body library \pmwd\ \citep{Li2022b},
which is an implementation of the differentiable cosmological simulation
with the adjoint method \citep{Li2022a}.

Here we briefly recap the equations that are closely related to our
extensions of the original framework, and introduce the notations of the
variables.
Following \cite{Li2022a}, we denote the particles' state with $\vz =
(\vx, \vp)^\intercal$, where $\vx$ and $\vp$ are the positions and
momenta of the particles, and their adjoint variables with $\vlambda =
(\vxi, \vpi)$.
We refer to the quantities that are directly related to the final
objective function $\cJ$ as the observable $\vo$.
In this work, our observables include multiple snapshots at different
times.
For the observation operator $O_{i}^{i-1}$ in the adjoint equations
(check Eq.(49) in \cite{Li2022a}) from time step $a_i$ back to
$a_{i-1}$, the adjoint variables are updated through
\begin{align}
  O_i^{i-1}: \qquad \vxi_{i-1} &:= \vxi_{i-1} +
              \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p \vx_{i-1}}, \nonumber\\
  \vpi_{i-1} &:= \vpi_{i-1} +
                          \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p \vp_{i-1}},
\end{align}
where the derivatives of $\vo$ with respect to $\vx_{i-1}$ or
$\vp_{i-1}$ would be non-zero only if $\vx_{i-1}$ or $\vp_{i-1}$ was
used in the generation of any snapshots in $\vo$.
The initial conditions of these adjoint variables are given by
\begin{align}
  \vxi_n &= \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p\vx_n}, \nonumber\\
  \vpi_n &= \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p\vp_n},
\end{align}
which are zero if no snapshot in $\vo$ depends on $(\vx_n, \vp_n)$, the
state of particles at the final time step.
The derivative of the objective function with respect to the observable,
i.e. $\p\cJ/\p\vo$, depends on our construction of the loss function on
the snapshots (\autoref{sec:loss}).
The VJPs of the observation operation, i.e. $\p\vo/\p\vx_n$ and
$\p\vo/\p\vp_n$, mainly involves the interpolation function
(\autoref{sec:snapobs}).
Both can be conveniently computed by AD.

The $N$-body time integration is carried out along discrete time steps
in terms of the scale factor $a$, and the state of particles at any
given time can be achieved using interpolation.
Given the particle displacements $\vs$ and velocities\footnote{This
velocity is related to the canonical momentum $\vp$ through $\vv =
\vp/(a^3H)$.} $\vv := \d\vs/\d a$ at two consecutive time steps $a_i$ and
$a_{i+1}$, we get the snapshot of particles in between at $a$ using the
cubic Hermite interpolation.
The interpolated particle displacements are given by
\begin{align}
  \vs(a) =\ &h_{00}(\alpha)\vs_i + h_{10}(\alpha)(a_{i+1} - a_i)\vv_i + \nonumber\\
           &h_{01}(\alpha)\vs_{i+1} + h_{11}(\alpha)(a_{i+1} - a_i)\vv_{i+1},
\end{align}
where $\alpha := (a - a_i)/(a_{i+1} - a_i)$ and the Hermite basis
functions read
\begin{align}
  h_{00}(\alpha) &= 2\alpha^3 - 3\alpha^2 + 1, \nonumber\\
  h_{10}(\alpha) &= \alpha^3 - 2\alpha^2 + \alpha, \nonumber\\
  h_{01}(\alpha) &= -2\alpha^3 + 3\alpha^2, \nonumber\\
  h_{11}(\alpha) &= \alpha^3 - \alpha^2.
\end{align}
The particle velocities are then given by the derivative $\vv(a) =
\d\vs(a) / \d a$.
The interpolation is a linear combination of the particle snapshots at
two consecutive time steps, which therefore can be done in either two
steps or just one step by carrying the particles from the previous step.
We find the two-steps scheme more convenient for the backward evolution
of the adjoint variables.
The interpolated snapshots at different times make the observable in
this work, and the VJPs of the interpolation functions are computed by
AD and put into the adjoint equations as mentioned in
\autoref{sec:pmwd}.

We sample 512 data points in the 9D space of configurations and
cosmological parameters using randomized quasi-Monte Carlo (RQMC) method
with scrambled Sobol sequence (\autoref{fig:sobol}).
We use the \texttt{scipy.stats.qmc} package \citep{SciPy} to generate
the Sobol sequence \citep{Sobol1967}, which uses the direction number
from \citet{JoeKuo2008} and the Owen scrambling \citep{Owen1998}.
We search among 65536 scrambling seeds to minimize the mixture
discrepancy (a uniformity measure) proposed in \citet{Zhou2013MD}.

Small numerical differences in initial conditions (ICs) could become
more significant with the $N$-body evolution.
We verified that the IC generation on GPUs is deterministic though
different from those on CPUs, therefore we generate ICs for both the
training data and the \pmwd\ model on GPUs.

We design the input features of the NNs in a way such that they could
properly fit and interpolate given our wide coverage of the
configurations and cosmological parameters.

For the displacement field, we adopt the mean squared error (MSE) loss
that proved effective in many previous works
\citep[e.g.,][]{HeEtAl2019, LiEtAl2021}:
%
\begin{equation}
\cL_\vs = \ln \frac{\sum_\vq \bigl[ \hat\vs(\vq) - \vs(\vq) \bigr]^2}
                   {\sum_{\vq'} s(\vq')^2},
\end{equation}
%
in which a particle originating from its Lagrangian position $\vq$ is
displaced by $\vs$ in the \GADGET\ snapshot and by $\hat\vs$ in the
prediction of \pmwd\ model in training.
Note that we have further taken the logarithm to combine it with the
other loss(es) below.
% Say that only s loss is not enough even though in principle it is,
% and cite DrewEtAl x2 too

Besides the mesh for the force evaluation, we also need meshes when
computing the power spectra of the density field for the loss function
as well as for visualization.

To perform further analyses of the trained NNs, including feature
importance analysis and symbolic regression fitting, we need to generate
samples of (input, output) data.
The input features of the NNs are constructed based on the RQMC
parameters in \autoref{tab:param}, the wavenumber $k$, and the scale
factor $a$.
Again, we use Sobol sequence to sample data points in this high
dimensional space.

\section*{TODO}
\begin{itemize}
\item

@@

  example}).

@@

@@
